Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-08-15 manuscript/usses_water_ms.tex
updates all model related stuff to ARK specification

Our main goal was to test whether the relationship between ANPP and soil
moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site
\citep{Blaisdell1958, Dalgleish2011, Adler2012}, we chose to use
cumulative volumetric water content from March through June as our
metric of soil moisture (hereafter referred to simply as `soil
moisture'). To achieve this goal, we fit a multi-level random intercept
and random slope regression with log(ANPP) as the response variable and
soil moisture as the sole predictor. We log-transformed ANPP to account
for heteroscedasticity. We fit the model under a Bayesian framework,
allowing us to test for treatment differences by comparing the posterior
distributions of the treatment-level coefficients
\citep[e.g.,][]{Tredennick2013}. Both log(ANPP) and soil moisture were
standardized to have mean 0 and unit variance before fitting the model
{[}i.e., \((x_i - \bar{x})/\sigma_x\){]}.

@@

\begin{align}
\mu_{j(k(t))} &= \beta_{0,j(k)} + \beta_{1,j(k)}x_{k(t)} + \gamma_t, \\
y_{j(k(t))} &\sim \text{Normal} \left(\mu_{j(k(t))}, \sigma^2_{k} \right),
\end{align}

\noindent{}where \(\mu_{j(k(t))}\) is the deterministic prediction from
the regression model for plot \emph{j} associated with treatment
\emph{k} in year \emph{t}, \(\beta_{0,j(k)}\) is the intercept for plot
\emph{j} associated with treatment \emph{k}, \(\beta_{1,j(k)}\) is the
slope term for the effect of soil moisture for plot \emph{j} associated
with treatment \emph{k}, \(\gamma_t\) is the intercept offset for year
\emph{t}, and \(\sigma^2_k\) is the process variance for treatment
\emph{k}. Data include the standardized log(ANPP) observations
\(\left( y_{j(k(t))} \right)\) and soil moisture (\(x_{k(t)}\)).

The intercept and slope terms are modeled hierarchically to account for
the non-independence of observations across years within plots and to
allow us to test the hypothesis that our treatments alter the ANPP-soil
moisture relationship. Thus, our multi-level model has three grouping
levels for coefficients, representing the nested structure of the data:
(i) overall coefficients, (ii) treatment coefficients, and (iii) plot
coefficients. Each subsequent level is drawn from the distribution of
coefficients at the previous level, meaning plot-level coefficients are
drawn from treatment-level coefficients, which are drawn from overall
coefficients. We also include a covariance structure among the intercept
and slope at each level. Formally, our hierarchical structure is as
follows, where we drop the intercept (0) and slope (1) subscripts and
instead refer to a vector of coefficients, \(\boldsymbol{\beta}\):

@@

\begin{align}
\boldsymbol{\beta}_{j(k)} &\sim \text{MVN} \left( \boldsymbol{\beta}_k, \boldsymbol{\Sigma}(k) \right), \\
\boldsymbol{\beta}_{k} &\sim \text{MVN} \left( \boldsymbol{\beta}, \boldsymbol{\Sigma}  \right), \\
\boldsymbol{\beta} &\sim \text{Normal} \left( 0, 1 \right),
\end{align}

\noindent{}where \(\boldsymbol{\beta}_{j(k)}\) is the vector of
regression coefficients (intercept and slope) for plot \emph{j}
associated with treatment \emph{k}, \(\boldsymbol{\beta}_{k}\) is the
vector of coefficients for each treatment, and \(\boldsymbol{\beta}\) is
the vector of overall coefficients. The plot- and treatment-level
coefficients are drawn from multivariate normal distributions with
covariance matrix \(\boldsymbol{\Sigma}\). For the plot-level
coefficients, each treatment has its own variance-covariance matrix
(i.e., \(\boldsymbol{\Sigma}(k)\)). The overall coefficients are drawn
from a normal prior with mean 0 and standard deviation 1. The random
year effects (\(\boldsymbol{\gamma}\)) are drawn from a normal prior
with mean 0 and standard deviation \(\sigma_{\text{year}}\), which was
drawn from a weibull distribution. A full description of model is in
Appendix 2.

@@

\begin{table}[ht]
\centering
\caption{Results from statistical tests for clustering and dispersion of community composition among precipitation treatments. `adonis` tests whether treatments form unique clusters in multidimensial space; `betadisper` tests whether treatments have similar dispersion. For both tests, \emph{P} values greater than 0.05 indicate there is no support that the treatments differ.} 
\begingroup\normalsize
\begin{tabular}{rlrrrr}
  \hline
Year & Test & n & d.f. & \emph{F} & \emph{P} \\ 
  \hline
2011 & adonis &  21 &   2 & 1.02 & 0.42 \\ 
  2011 & betadisper &  21 &   2 & 2.23 & 0.14 \\ 
  2012 & adonis &  22 &   2 & 1.10 & 0.34 \\ 
  2012 & betadisper &  22 &   2 & 0.21 & 0.81 \\ 
  2013 & adonis &  22 &   2 & 1.23 & 0.14 \\ 
  2013 & betadisper &  22 &   2 & 0.28 & 0.76 \\ 
  2014 & adonis &  22 &   2 & 0.95 & 0.54 \\ 
  2014 & betadisper &  22 &   2 & 0.35 & 0.71 \\ 
  2015 & adonis &  21 &   2 & 1.05 & 0.40 \\ 
  2015 & betadisper &  21 &   2 & 3.01 & 0.07 \\ 
  2016 & adonis &  21 &   2 & 1.07 & 0.33 \\ 
  2016 & betadisper &  21 &   2 & 0.50 & 0.62 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/glmm_main_results.png}
  \caption{Results from the generalized linear mixed effects model. (A) Posterior distributions of treatment-level parameters ('Intercept' and the effect of 'Soil Moisture'). Kernel bandwidths of posterior densities were adjusted by a factor of 5 for visual clarity. (B) Scatterplot of the data and model estimates shown a solid lines. Model estimates come from treatment level coefficients (colored lines) and the overall mean coefficients (heavy black line). Note that we show log(ANPP) on the \emph{y}-axis of panel B; this same plot can be seen on the arithmetic scale in supporting material Fig. A2-1.}
\end{figure}

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/glmm_yeardiffs.png}
  \caption{Posterior distributions of random year effects (intercept offsets). Kernel bandwidths of posterior densities were adjusted by a factor of 5 for visual clarity.}
\end{figure}

>>>>

Our main goal was to test whether the relationship between ANPP and soil
moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site
\citep{Blaisdell1958, Dalgleish2011, Adler2012}, we chose to use
cumulative volumetric water content from March through June as our
metric of soil moisture (hereafter referred to simply as `soil
moisture'). To achieve this goal, we fit a hierarchical regression model
with log(ANPP) as the response variable and soil moisture as the sole
predictor. We log-transformed ANPP to account for heteroscedasticity.
Both log(ANPP) and soil moisture were standardized to have mean 0 and
unit variance before fitting the model {[}i.e.,
\((x_i - \bar{x})/\sigma_x\){]}.

@@

\begin{align}
\mu_{i} &= \boldsymbol{\beta}\textbf{x}_i + \boldsymbol{\gamma}_{j(i)}\textbf{z}_i + \eta_t, \\
\textbf{y} &\sim \text{Normal} \left(\boldsymbol{\mu}, \sigma^2 \right),
\end{align}

\noindent{}where \(\mu_{i}\) is the deterministic prediction from the
regression model for observation \emph{i}, which is associated with plot
\emph{j} and treatment year \emph{t}. \(\boldsymbol{\beta}\) is the
vector of coefficients for the fixed effects in the design matrix
\(\textbf{X}\). Each row of the design matrix represents a single
observation (\(\textbf{x}_i\)) and is a vector with the following
elements: 1 for the intercept, a binary 0 or 1 if the treatment is
``drought'', a binary 0 or 1 if the treatment is ``irrigation'', the
scaled value of volumetric water content (VWC), binary ``drought'' value
times VWC, and binary ``irrigation'' value times VWC. Thus, our model
treats ``control'' observations as the main treatment and then estimates
intercept and slope offsets for the ``drought'' and ``irrigation''
treatments. In reference to our model, the hypotheses we wish to test
are: (1) the coefficient for drought\(\times\)VWC is positive and
different from zero, and (2) the coefficient for irrigation\(\times\)VWC
is negative and different from zero

To account for the fact that observations within plots and years are not
independent, we include three random effects. Specifically, we include
plot-specific offsets (\(\boldsymbol{\gamma}\)) for the intercept and
slope terms and year-specific intercept offsets (\(\eta_t\)). The
covariate vector \(\textbf{z}_i\) for each observation \emph{i} has two
elements: a 1 for the intercept and the scaled value of VWC for that
plot and year. The plot-specific coefficients are modeled
hierarchically, where plot level coefficients are drawn from a
multivariate normal distribution with mean 0 and a variance-covariance
strucutre that allows the intercept and slope terms to be correlated:

@@

\begin{align}
\boldsymbol{\gamma}_{j(i)} &\sim \text{MVN} \left( 0, \Sigma  \right),
\end{align}

\noindent{}where \(\Sigma\) is the variance-covariance matrix and
\emph{j(i)} reads as ``plot \emph{j} associated with observation
\emph{i}''. The random year effects (\(\boldsymbol{\eta}\)) are drawn
from a normal prior with mean 0 and standard deviation
\(\sigma_{\text{year}}\), which was drawn from a half-cauchy
distribution. A full description of our model is in Appendix 2.

@@

\begin{table}[ht]
\centering
\caption{Summary statistics from the posterior distributions of coefficients for each treatment.} 
\begingroup\normalsize
\begin{tabular}{llrrrr}
  \hline
Coefficient & Treatment & Posterior Mean & Posterior Median & Lower 95\% BCI & Upper 95\% BCI \\ 
  \hline
Intercept & Control & 0.02 & 0.03 & -1.06 & 1.01 \\ 
  Intercept & Drought & -0.04 & -0.04 & -1.39 & 1.21 \\ 
  Intercept & Irrigation & -0.13 & -0.11 & -1.35 & 1.09 \\ 
  Slope & Control & 0.63 & 0.64 & -0.33 & 1.61 \\ 
  Slope & Drought & 0.46 & 0.47 & -0.59 & 1.52 \\ 
  Slope & Irrigation & 0.26 & 0.26 & -0.60 & 1.12 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}\begin{table}[ht]
\centering
\caption{Results from statistical tests for clustering and dispersion of community composition among precipitation treatments. `adonis' tests whether treatments form unique clusters in multidimensial space; `betadisper' tests whether treatments have similar dispersion. For both tests, \emph{P} values greater than 0.05 indicate there is no support that the treatments differ.} 
\begingroup\normalsize
\begin{tabular}{rlrrrr}
  \hline
Year & Test & n & d.f. & \emph{F} & \emph{P} \\ 
  \hline
2011 & adonis &  21 &   2 & 1.02 & 0.42 \\ 
  2011 & betadisper &  21 &   2 & 2.23 & 0.14 \\ 
  2012 & adonis &  22 &   2 & 1.10 & 0.34 \\ 
  2012 & betadisper &  22 &   2 & 0.21 & 0.81 \\ 
  2013 & adonis &  22 &   2 & 1.23 & 0.14 \\ 
  2013 & betadisper &  22 &   2 & 0.28 & 0.76 \\ 
  2014 & adonis &  22 &   2 & 0.95 & 0.54 \\ 
  2014 & betadisper &  22 &   2 & 0.35 & 0.71 \\ 
  2015 & adonis &  21 &   2 & 1.05 & 0.40 \\ 
  2015 & betadisper &  21 &   2 & 3.01 & 0.07 \\ 
  2016 & adonis &  21 &   2 & 1.07 & 0.33 \\ 
  2016 & betadisper &  21 &   2 & 0.50 & 0.62 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}

@@

\begin{figure}[!ht]
  \centering
      \includegraphics[width=5in]{../figures/glmm_main_results.png}
  \caption{Results from the generalized linear mixed effects model. (A) Posterior distributions of the intercept and slope offsets for the drought and irrigation treatments. Offsets indicate the amount to which the coefficients for drought or irrigation treatments differ from the control treatment estimates. Probabilities (``Pr $<$ 0 ='') for each distribution indicate the probability that coefficient is less than zero. Probabilities greater than 0.95 indicate strong support for the coefficient being less than zero. We only show the one-tailed probability for the value being less than zero because the median of each distribution is less than zero. Kernel bandwidths of posterior densities were adjusted by a factor of 5 for visual clarity. (B) Scatterplot of the data and model estimates shown a solid lines. Model estimates come from treatment level coefficients (colored lines). Note that we show log(ANPP) on the \emph{y}-axis of panel B; this same plot can be seen on the arithmetic scale in supporting material Fig. A2-1.}
\end{figure}
